---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Load libraries and set global defaults.

```{r,warning=FALSE}
##Load any required libraries and set any global defaults
if (!require("knitr"))
        {
                install.packages("knitr")
        }

if (!require("ggplot2"))
        {
                install.packages("ggplot2")
        }

if (!require("Hmisc"))
        {
                install.packages("Hmisc")
        }

# Turn off scientific notations for numbers
options(scipen = 999)  

#Set globals
opts_chunk$set(fig.width=7, fig.height=7, warnings=FALSE)  
Sys.setlocale("LC_ALL", "English") 

#Set working directory to data directory
setwd('C:/Users/bryan_000/Documents/GitHub/Data/')
```

## Loading and preprocessing the data

The variables included in this data set are:

        steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

        date: The date on which the measurement was taken in YYYY-MM-DD format

        interval: Identifier for the 5-minute interval in which measurement was taken
```{r}
if (!file.exists("activity.csv"))
 {
        data.url <- 'http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
        zip.file <- 'activity.zip'
        download.file(data.url, zip.file) 
        unzip(zip.file)
 }

##Read the data
data <- read.table("activity.csv", header = TRUE, sep =",")
```

The data set is stored in a comma-separated-value (CSV) file.

        There are a total of `r nrow(data)` observations in this data set.  
        There are a total of `r sum(is.na(data$steps))` days/intervals with missing values in this data set.

## What is mean total number of steps taken per day?

For this part of the assignment, missing values in the data set will be ignored. For step 1, I prepared a Histogram of the total number of steps taken each day
```{r}
totalsteps <- with(data, rowsum(steps, date))

#plot the histogram
hist(totalsteps, breaks=8, col="blue", main="Total Steps Histogram", xlab="Steps")
rug(totalsteps)
```

For the second part, I calculated the mean and median total number of steps taken per day (ignoring missing values).  The mean number of steps is `r round(mean(totalsteps,na.rm=TRUE),2)` and the median number of steps is `r median(totalsteps, na.rm=TRUE)`.

## What is the average daily activity pattern?


I calculated the aggregation of steps by intervals of 5-minutes and saved them in a data frame called stepsperint. I then plotted the average number of steps taken (averaged across all days) versus the 5-minute intervals.
```{r}
## Calcutae mean steps per Interval for plotting
stepsperint<- aggregate(data$steps,by = list(interval = data$interval), FUN=mean, na.rm="TRUE")

##Plot 
with(stepsperint, plot(interval, x, type="l", col="red"))
```

Next, I computed the 5-minute interval with the containing the maximum number of steps.
```{r}
maxnumsteps <-stepsperint[which.max(stepsperint[,2]),]
```

On average, the interval `r maxnumsteps$interval` was the 5-minute interval which contained the maximum number of steps `r round(maxnumsteps$x,0)`.

## Imputing missing values

There are many days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.  

```{r}
originalValue <- complete.cases(data)  
nMissing <- length(originalValue[originalValue==FALSE])     # number of records with NA  
nComplete <- length(originalValue[originalValue==TRUE])     # number of complete records

title="Missing vs. Complete Cases"  
barplot(table(originalValue),main=title,xaxt='n', col="bisque3")# render Complete Cases barplot  
axis(side=1,at=c(.7,1.9),labels=c("Missing","Complete"),tick=FALSE)    # render axis  
text(.7,0,labels=nMissing, pos=3)                                  # label the NA's bar  
text(1.9,0,labels=nComplete, pos=3)   
```

This data set had `r sum(is.na(data$steps))' days/intervals with missing values.  I copied the original data set to a new data set and added a column "imputed_steps" and filled the missing values with the mean value.

```{r}
impdata <-data
impdata$imputed_steps<-with(impdata, impute(steps,mean))
```

The data set with the missing steps imputed had `r  sum(is.na(impdata$imputed_steps))` days/intervals with missing values.

Now, using the data set with the missing steps imputed, I made a histogram of the total number of steps taken each day and calculated the mean and median total number of steps.
```{r}
imptotalsteps <- with(impdata, rowsum(imputed_steps, date))

#plot the histogram
hist(imptotalsteps, breaks=8, col="blue", main="Total Steps Histogram After NA Imputed", xlab="Steps")
rug(imptotalsteps)
```

###Differences

The mean number of steps calculated with the imputed values was `r round(mean(imptotalsteps),2)` and the median number of steps was `r round(median(imptotalsteps),2)`.  The mean and median previously calculated by ignoring the NA values were `r round(mean(totalsteps,na.rm=TRUE),2)` and `r round(median(totalsteps,na.rm=TRUE),2)` respectively.   

###Impact of imputing missing data

The mean was unchanged by imputing the missing values using the mean value for the 5 minute interval.  The median increased slightly to match the mean. As expected, the frequencies of each bin were higher with the imputed missing data.

## Are there differences in activity patterns between weekdays and weekends?

I conducted this analysis using the data set with missing values imputed.  I made a plots containing average number of steps taken on weekdays and weekends.
```{r}
impdata$week <- factor(weekdays(as.Date(data$date)) %in% c("Saturday", "Sunday"), c("FALSE", "TRUE"), c("weekdays", "weekend"))

stepsperint<- aggregate(impdata$imputed_steps, by = list(interval=impdata$interval, week=impdata$week), FUN = mean, na.rm = TRUE)

##Plot the output
ggplot(stepsperint, aes(x = interval, y = x)) + geom_line() + facet_wrap(~week, ncol=1)
```

The graph above that activity on the weekday has the greatest peak from all steps intervals. The weekends activities has more peaks over a hundred than the weekday activities. 